### 2012-06-12 trunk/ijcis-pos-model/sources/temporaldatabases.tex
My contributions so far...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Time
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\label{subsec:time-in-databases}Time in Databases}
The concept of time has been studied in databases for a long time. A true standard for adding temporal aspects to relational databases does not exist, but there is a consensus in the literature \cite{Dyreson1994} on what is called a \emph{temporal database}: a temporal database is a database dealing with some aspects of time in its schema.
In a temporal DBMS, a \textbf{chronon} is the shortest duration of time supported by the system. In temporal databases, some temporal attributes can be managed without treating the attribute differently from non-temporal attributes. The time described by such an attribute is called \textbf{user defined time} (\emph{UDT}). In addition to UDT, the following types of time can be discerned in a temporal database, all of which are handled exceptionally by the DBMS:

\begin{itemize}
	\item
	\textbf{Transaction time} (\emph{TT}) \cite{Rowe1987},\cite{Jensen1991} denotes the time when the fact (object) is stored in the database. It is usually append-only: as the past can not be changed, TT can not be changed neither. Furthermore, at the moment of insertion, a TT can be neither in the past nor in the future.
	\item
	\textbf{Valid time} (\emph{VT}) \cite{Jensen1994},\cite{Sarda1990},\cite{McKenzie1981} denotes the time when the fact (object) is true in the modelled reality. A fuzzy extension has been proposed by \cite{Garrido2009}. 
%	\item
%	User defined time: It is an uninterpreted attribute. The domain is date/time. The query language has no special support for it.
	\item
	\textbf{Decision time} (\emph{DT}) \cite{Nascimento1995},\cite{Chakravarthy1994},\cite{Etzion1992},\cite{Ozsoyoglu1995} denotes the time when an event was decided to happen. 
	\end{itemize}

	E.g., consider a database containing employee contract descriptions. The time when the employee's contract is valid, represented by an interval, is VT. The time when the employee's contract is stored in the database is the TT. The time when the decision for hiring this employee was made is the DT.

% is a non-decomposable unit of time.  it  There are two ways to represent a chronon: as a point or as an interval \cite{655777}.

When working with these time concepts, the Data Manipulation Language (\emph{DML}, which is part of the standard database querying language SQL) is extended to deal with possible temporal inconsistencies within the data and to handle more complex (temporal) queries. 
%Next to these concepts, also \textbf{user defined time} (\emph{UDT}) is discerned. UDT is an uninterpreted attribute in the date/time domain. This means that the attribute uses the date/time domain, but the database model does not treat the attribute differently from non-temporal attributes.
Depending on the time managed, a database is classified as either a \textbf{Valid Time Database} (\emph{VTDB}), a \textbf{Transaction Time Database} (\emph{TTDB}), a \textbf{bi-temporal database} (both valid and transaction time are managed) or a \textbf{tri-temporal} or \emph{multitemporal} database (valid time, transaction time and decision time are managed).

%\subsection{Temporal Elements}
%%	
%	In order to define a temporal database model, some basic elements should be defined \cite{Dyreson:1994:CGT:181550.181560}:
%	\begin{itemize}
%	\item A	\textbf{chronon} is a non-decomposable unit of time. In a temporal DBMS, it is the shortest duration of time supported by the system. There are two ways to represent a chronon: as a point or as an interval \cite{655777}.
%	\item
%	\emph{Event}: An instant of time. Usually, an event is said to be occur during time $t$ if it occurs during the chronon represented by $t$.
%	\item
%	\emph{Interval}: The time between two events. The representation is very often a set of contiguous chronons.
	%\item
	%\emph{Span}: A directed duration of time. A duration is an amount of time with known  lenght but no specific starting or ending chronons. The span can be either positive, denoting a forward motion of time or negative, denoting a backward motion of time.
%	\item A	\textbf{timestamp} is a time value associated with some object in the database.

%	\item The	\textbf{lifespan} (of a database object)is the time over which it is defined. The lifespan of a valid time object denotes the time when the corresponding object exists. The lifespan of a transaction time object is the value of the timestamp.

%	Depending on the type of time the meaning is different:
%		\begin{itemize}
%		\item
%		Lifespan of a valid time object: Refers the time when the corresponding object exists.	
%		\item
%		Lifespan of a transaction time object: The transaction time of a database object refers when the object is stored in the database. The lifespan is the value of the timestamp.		
%		\end{itemize}
%	\end{itemize}

%	In the temporal database thesaurus, \emph{'Snapshot'} is the word for non-temporal matters. As a temporal database is a generalization for relational databases, an \emph{snapshot database} is a relational database. Furthermore, a \emph{snapshot relation} is a relation incorporating neither valid nor transaction time.

%\subsection{Main issues when dealing with time}
%Among others, the following problems are present when dealing with time in a database:

%	\begin{itemize}	
%	\item \textbf{Granularity} denotes a partition on the set of chronons. The conversion between several granularities is studied in \cite{Lin97efficientconversion}. Granularity is the base of the temporal model in \cite{Cru97}. An object-oriented implementation is in \cite{624013}.
%	\item	To ensure \textbf{consistency}, temporal databases usually redefine the primary key of a relation. The new primary key takes into account the presence of the time. In order to keep the consistency, the DML is re-defined. For example, in a VTDB, the temporal update sentence is usually composed by an update sentence (closing the old row) plus an insert sentence (creating the new row).
%	\end{itemize}
%Guy's suggestion: instead of imprecision, use imperfection which is more general.
\subsubsection{Imperfection and time}
Representing imprecision and its semantics when dealing with time has been studied for a long time. Several proposals for representing and computing imprecise time indications can be found in \cite{DeCaluwe1997} and \cite{DeTre1997}. Also, the changes between several granularities can be seen as a source of imprecision \cite{Devos1998}.

In the proposal section we will consider two kinds of imprecision:
\begin{itemize}
\item \textbf{Uncertainty in the database} denotes the uncertainty that arises when the knowledge about the temporal data in the database is uncertain. E.g., a database record shows that \emph{`The car is in the garage around April.'}
 \item \textbf{Imprecision in the query specification} denotes the imprecision in the specification of temporal criteria by the user, when querying. E.g., \emph{`The user wants to obtain a car which is red and which is in the garage around April.'}
\end{itemize}

\subsubsection{Representation}
Several proposals for managing uncertain time in a database exist. Some proposals work with rough sets \cite{Qiang2009}, other proposals rely on possibility distributions for representing uncertainty in time \cite{Dyreson1998},\cite{Garrido2009}, \cite{Galindo2001}. In order to compare temporal possibility distributions, extensions of the classical Allen's operators \cite{Allen1983} are defined in \cite{Ohlbach2004}, \cite{Nagypal2003},\cite{Dubois2003a} and \cite{Schockaert2008}
%In the proposal section, we will follow the representation by means of possibility distributions, in order to work with both satisfaction and dissatisfaction degrees. Also, in order to work properly with fuzzy operators, the underlying domain should be numeric. 
% Take a look into this:
%In this paper, the representation for the dates will follow the Julian Day Number (JDN) representation \cite{Dir96}.

%If the starting points and/or the end points of the interval representing the time are not known precisely, it is easy to fuzzify them, using, e.g., two triangular possibility distributions.

In order to deal with uncertainty in time intervals, several proposals are made. Here, two approaches are introduced: the first one, based on \emph{Fuzzy Validity Period}~\cite{Garrido2009} and the second one based on \emph{Possibilistic Valid-time Period}~\cite{JoseEnriquePons2012}.

\begin{definition}
A \textbf{Fuzzy Validity Period} (\emph{FVP}) is defined as a fuzzy time interval specifying when an object is valid. A fuzzy time interval is then the fuzzification of a crisp time interval.
\end{definition}
Several options to transform possibility distributions corresponding to the fuzzy starting point and the fuzzy end point into one consistent FVP exist \cite{Garrido2009}, e.g (Fig. \ref{fig:fuzzy-validity-period}):
\begin{itemize}
\item The \textbf{convex hull} approach is the most intuitive approach. The resulting FVP is the convex hull of the union of both fuzzy sets.
\item The \textbf{uncertainty preserving} approach is less intuitive but more realistic. The amount of uncertainty is maintained at the edges of the possibility distribution representing the FVP \cite{Garrido2009}.
\end{itemize}

%%%%%%
% FUZZY VALIDITY PERIOD
%%%%%%
\vspace*{13pt}
\begin{center}
{
\includegraphics[scale=0.25]{./graphs/comparisoncv.pdf}

}
\end{center}
%\centerline{ \psfig{file=./graphs/Y-time-point.eps}}
\vspace*{10pt}
\fcaption{\label{fig:fuzzy-validity-period}Transformation to obtain the FVP. The top graph shows the two triangular possibility distributions. The middle graph shows the convex hull validity period, the bottom one shows the result of the second transformation, which maintains the imprecision.}
%  \label{fig:fuzzy-validity-period}
\vspace*{13pt}

\begin{definition}
A \textbf{Possibilistic Valid-time Period} (\emph{PVP}) is an ill-known interval in time specifying when an object is valid.
\end{definition}
A PVP is an ill-known interval in the sense of definition \ref{def;possibilistic-variable}, section \ref{subsec:possibilistic-variables}. Note that this representation is \emph{disjunctive}: the PVP represents only one crisp time interval, but that for some reason it is (partially) unknown.

The ill-known interval approach has many advantages from the representation by the FVP as demonstrated in \cite{Pons2011},\cite{Pons2012}. Table \ref{tbl:comparative-pvp-fvp} is a comparative between PVP and FVP. The following list defines the items in the comparative:

\begin{enumerate}
\item Domain: The domain of the possibility distribution modelled by the approach.
\item Implementation of relationships: How to implement a relationship.
\item Allen's relations: Are the Allen's relations defined?
\item Storage: The way the data is stored in the database.
\item Possibility measures: Does the framework provides always a possibility measure for any relation between the temporal elements?
\item Necessity measures:  Does the framework provides always a necessity measure for any relation between the temporal elements?
\end{enumerate}

%%
%% comparison table between pvp and fvp.
%%
\vglue13pt
%\begin{table}[htbp]
\tcap{Comparative PVP vs FVP}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c p{2cm} p{2cm}}\\
Item & PVP & FVP\\
\hline
(1) & $\Pow(\R)$ & $\R$ \\
(2) & Ill-known constraints. & Ad-hoc operators. \\
(3) & $\checkmark$ & - \\
(4) & Two distributions one for each endpoint. & Only one distribution. \\ 
(5) & $\checkmark$ & $\checkmark$ \\
(6) & $\checkmark$ & - \\
\hline\\
\end{tabular}
\label{tbl:comparative-pvp-fvp}
}

In the rest of the paper we will work only with PVP to represent valid-time intervals.

\subsection{Understanding Valid-time Databases}
This subsection is devoted to describe the behaviour of a crisp valid-time database. For the shake of simplicity, only the three main operations in the Data Manipulation Language \emph{DML} a.k.a \emph{CRUD} (CReate Update, Delete) operations are shown. Usually the DML operations in a temporal database are re-defined e.g. a typical update sentence in SQL could be re-defined by means of several insert and update sentences. Therefore, in order to be clear, these high level primitives in the DML for a valid-time database are usually noted as Insert, Modify and Delete. In the following subsections each primitive is defined and explained. Finally a illustrative example is given. For a more complete information on the behaviour of a bitemporal database, please refer to ~\cite{Jensen1994}.

\begin{definition}
Consider a relation $R$ in a temporal database, an object $A$ given by the attributes $\left(a_1, \ldots, a_n \right)$ and the crisp time interval $I$ given by $\left[a,\ b\right] : a,b \in \T, a \leq b$. A valid-time object $A$ in the relation $R$ is specified by:
\begin{equation}
\label{eq:rel-def}
\left( R, A , I \right)
\end{equation}
\end{definition}

In order to simplify the algorithms for the manipulation of data, some auxiliary functions are defined:

\begin{definition}
\label{def:close-a-crisp-interval-r}
Consider two crisp intervals $I= \left[a ,b \right]$ and $J= \left[c ,d \right]$, $c > a$. The CloseR$\left(I, J\right)$ function allows to close the right-open interval $I$ with respect to the first value $d$ in $J$:
\begin{align}
\mbox{CloseR} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } b \neq +\infty \\
I=\left[a, c \right[ & \mbox{ if } b = +\infty \wedge J > I
\end{cases}
\end{align}
\end{definition}

@@

Now it is possible to close the current version of an object by using \eqref{def:close-a-crisp-interval-r} and \eqref{eq:current-in-relation}

\begin{definition}
\label{def:close-current-version}
Consider an object $A$, the relation $R$ and a time interval $J$. Then, the function close-current$\left(R,A,J \right)$ closes any current version of the object $A$ if it exists and add the new version $\left(A, J \right)$.

\begin{eqnarray}
%\label{eq:close-current}
\text{Close-current} \left(R, A, J \right) =\\
\begin{cases}
\nonumber
\big \lbrace R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseR } \left(I, J\right) \right) \cup \left(A, J\right)\right \rbrace  \big \rbrace \\
\nonumber
\mbox{ if } I = \mbox{ Current } \left(R, A \right) \wedge J > \mbox{ CloseR } \left(I, J \right)   \\
\nonumber R , \text{ in any other case}
\end{cases}
\end{eqnarray}
\end{definition}

%It is also possible to append at the beggining by using \eqref{def:close-a-crisp-interval-l} and \eqref{def:first-in-relation}.
%\begin{definition}
%\label{def:append-first-version}
%Consider an object $A$, the relation $R$ and a time interval $J$. Then, the function append-first$\left(R,A,J \right)$ closes any right open version of the object $A$ if it exists and add the new version $\left(A, J \right)$.
%
%\begin{eqnarray}
%%\label{eq:close-current}
%\text{Append-first} \left(R, A, J \right) =\\
%\begin{cases}
%\nonumber
%\big \lbrace R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseL } \left(I, J\right) \right) \cup \left(A, J\right)\right \rbrace  \big \rbrace \\
%\nonumber
%\mbox{ if } I = \mbox{ First } \left(R, A \right) \wedge J < \mbox{ CloseL } \left(I, J \right)   \\
%\nonumber R , \text{ in any other case}
%\end{cases}
%\end{eqnarray}
%\end{definition}

\subsubsection{\label{subsubsec:insert}Insert}
The user wants to store an object $A$ which is valid in the relation $R$ during the time interval specified by $I = \left(a, b \right)$.
%
%
%The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used. 
There are two main cases when performing a create operation:
\begin{enumerate}
\item The object $A$ was never in the relation $R$: The object is added with the valid-time indicated by the crisp interval $I$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Time
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Time in databases: temporal databases: the crisp stuff. General things.
\subsection{\label{subsec:time-in-databases}Time in Databases}
The concept of adding time to databases has been profoundly studied~\cite{Darwen1998},~\cite{Etzion1998},~\cite{Etzion1998b},~\cite{Gal1998},~\cite{Jensen1999},~\cite{Lomet2006},~\cite{Nascimento1995},~\cite{Sarda1990},~\cite{Snodgrass1998},~\cite{Toman1998},~\cite{Wu1998}. A true standard for adding temporal aspects to relational databases does not exist, but there is a consensus on what is called a \emph{temporal database}: a temporal database is a database that supports some aspects of time~\cite{Dyreson1994}. Thus, a temporal database will contain data representing \emph{temporal values}. Notions of time, such as instants~\cite{Dyreson1994}, time intervals~\cite{Dyreson1994} or temporal elements~\cite{Dyreson1994} are generally called `temporal values' in the context of temporal databases. Based on their interpretation and modelling purpose, time notions and the corresponding times can be classified into four types. Of these types, \emph{user-defined time} comprises time indications of which the modelling has no impact on the consistency of the temporal database. The data representing these temporal values are not handled differently from data of non-temporal attributes (by the temporal DBMS)~\cite{Dyreson1994}. The other types contain time indications typically represented by data which are handled exceptionally by the temporal DBMS. The definitions and explanations of these types can be found in~\cite{Dyreson1994} and~\cite{Nascimento1995} and more information can be found in~\cite{Jensen1991},~\cite{Snodgrass1984} and~\cite{Nascimento1995}.

\begin{itemize}
	\item \emph{Valid time}: The \emph{valid-time} (VT) of a fact is the time when the fact is true in the modeled reality~\cite{Dyreson1994}.
	\item \emph{Transaction time}: A database fact is stored in a database at some point in time, and after it is stored, it is current until logically deleted. The \emph{transaction-time} (TT) of a database fact is the time when the fact is current in the database and may be retrieved~\cite{Dyreson1994}.
	\item \emph{Decision time}: The \emph{Decision time} (DT) of an event is the time when an event was decided to happen~\cite{Nascimento1995}.
\end{itemize}

An example: consider a database containing data representing the capacities of various aspects of employee contracts. The time during which an employee's contract is valid and the employee is hired, will then be VT. The moment on which an employee's contract is stored in the database, will then be TT. The time when the decision for hiring this employee was made, will then be DT.

Depending on the time represented in the database, a database is classified as either a \emph{valid-time database} (VTDB), a \emph{transaction-time database} (TTDB), a \emph{decision-time database} (DTDB), a \emph{bitemporal database} (both valid-time indications and transaction-time indications are represented), or a \emph{tritemporal} or \emph{multitemporal database} (valid-time indications, transaction-time indications and decision-time indications are represented).

%Now in the context of a relational database: what is a VT/TT/... relation? We will use VT relations, thus we should explain what they are.
In the context of relational databases, more specific definitions exist. For example, a \emph{valid-time relational database} is a relational database containing one or more \emph{valid-time relations}, a \emph{transaction-time relational database} is a relational database containing one or more \emph{transaction-time relations} and a \emph{bitemporal relational database} is a relational database containing one or more \emph{bitemporal relations}~\cite{Dyreson1994}. A valid-time relation is then a relation with exactly one system-supported VT, a transaction-time relation is a relation with exactly one system-supported TT and a bitemporal relation is a relation with exactly one system-supported VT and exactly one system-supported TT~\cite{Dyreson1994}. The attributes used in a valid-time relation to describe VT are called \emph{valid-time attributes}. The attributes used in a transaction-time relation to describe TT are called \emph{transaction-time attributes}. Any attribute unrelated to VT, TT or DT is usually called a \emph{non-temporal attribute}.

%In a temporal database system, a \emph{chronon} is the shortest duration of time supported by the system. % is a non-decomposable unit of time.  it  There are two ways to represent a chronon: as a point or as an interval \cite{655777}. %WE NEVER USE CHRONONS IN THIS PAPER!
Typically, when time modelling is supported by a database system, some behavioral aspects of the database model are extended or enhanced and some extra behavioral aspects are introduced, to deal with possible temporal inconsistencies and to allow more complex (temporal) queries. The presented work only deals with valid-time relational databases. In the presented work, a general extended/enhanced version of the relational database model is used. This extended version is based on several works, mainly the work presented in~\cite{Jensen1994}. In subsection~\ref{subsubsec:Understanding-valid-time-databases}, the extensions to the relational database model used in this work are presented on a logical level, but first, in subsection~\ref{example-situation}, an example situation is presented, which will be used throughout this paper to explain the various concepts treated in light of the presented work.

%Thus, in the presented work, an extended version of the Data Manipulation Language DML, which is part of the standard database querying language SQL, is used. This extended version is based on several works, mainly the work presented in~\cite{Jensen1994}. In subsection~\ref{subsubsec:Understanding-valid-time-databases}, the extensions to the DML used in this work are presented, but first, in subsection~\ref{example-situation}, an example situation is presented, which will be used throughout this paper to explain the various concepts treated in light of the presented work.

\subsection{\label{example-situation}Example Situation}
Consider a valid-time relation modelling the absence through loaning of books in a library. The relation contains 4 attributes: `ID', `Title', `Start Loan' and `End Loan'. Every physical book is given a unique identifier. This identifier is represented by a number and described by attribute `ID'. Attribute `Title' describes the title of a book. Attributes `Start Loan' and `End Loan' describe the starting date, respectively ending date on which a valid-time interval starts, respectively ends. This time interval is the time interval during which the book, represented by the value of the tuple for the `ID' attribute, was loaned to a customer. Table \ref{tbl:library-sample} visualises this relation with a few tuples in the relation extention.

\tcap{\label{tbl:library-sample}Example library relation.}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Title  & Start Loan & End Loan \\
\hline
3 & Dracula & 15/3/2012 & 30/3/2012 \\
4 & Frankenstein & 25/3/2012 & 4/4/2012 \\
5 & Harry Potter & 18/3/2012 & 2/4/2012\\
3 & Dracula & 4/4/2012 & UC \\
\hline\\
\end{tabular}
}

%Valid-time databases and relations: general things AND an explanation about the exact things we will use in the rest of the paper. (+ vocabulary/concept explanation). Also explain the working of crisp valid-time databases/relations.
\subsection{\label{subsubsec:Understanding-valid-time-databases}General Valid-time Databases}
This subsection is devoted to describing the general behavior of general valid-time relational databases. This behavior will be described on a logical level, by (logically) defining and describing general forms of the four basic functions typically present in any relational database model. These functions are `create', `read', `update' and `delete' and will be denoted `the CRUD functions'. Usually, the CRUD functions in a valid-time relational database model are defined using the CRUD functions in a non-temporal relational database model. For example, the update function in a valid-time relational database model will typically be defined using several non-temporal create and update functions. In the presented work, the CRUD functions in the non-temporal relational database model will be given a subscript `NT', while the CRUD functions in the valid-time relational database model will be given a subscript `VT'.

Before defining and describing the CRUD functions used in the presented work, some concepts should be described and defined and the corresponding terminology and denotation style of these concepts should be determined.

%For the sake of simplicity, only the three main operations in the Data Manipulation Language \emph{DML} a.k.a \emph{CRUD} (CReate Update, Delete) operations are shown. Usually the DML operations in a temporal database are re-defined e.g. a typical update sentence in SQL could be re-defined by means of several insert and update sentences. Therefore, in order to be clear, these high level primitives in the DML for a valid-time database are usually noted as Insert, Modify and Delete. In the following subsections each primitive is defined and explained. Finally a illustrative example is given. For a more complete information on the behaviour of a bitemporal database, please refer to ~\cite{Jensen1994}.

Every valid-time relation has one VT. In the presented work, indications of VT which are represented in a valid-time relation are always time intervals. In the presented work, such a valid-time interval is represented in the valid-time relation using two instants~\cite{Dyreson1994}: a starting instant and an ending instant.

Consider a valid-time relation $R$ with attribute tuple $\left(A_{0},\ A_{1},\ A_{2},\ \ldots,\ A_{n},\ Start,\ End\right)$, where every $A_{i}$, $i \in \Nat$ and $i \geq 1$ and $i \leq n$ and $n \in \Nat_{0}$, denotes a non-temporal attribute and where $Start$ and $End$ denote temporal attributes, respectively describing starting and ending instants. Now consider an entity $E$ of which the properties are described by the aforementioned attributes $A_{i}$ and a time interval $I$ defined by instants $I_{s} \text{ and } I_{e}$ where $I_{s}$ and $I_{e}$ denote instants. Let the values representing these instants be denoted $s \in \T$ and $e \in \T$ respectively, $\T$ being the domain of the temporal attributes and let the representation of the time interval I be denoted $\left[s,\ e\right]$. Let the value representing the state of $E$ for attribute $A_{i}$ during time interval $I$ be denoted $a_{i}$, for every $i \in \Nat$ for which $i \geq 1$, $i \leq n$, and let the n+1-tuple $\left(a_{0},\ a_{1},\ a_{2},\ \ldots,\ a_{n}\right)$ be denoted $a$. The denotation

\begin{equation}
\label{eq:def-state}
\left(a,\ \left[s,\ e\right]\right)
\end{equation}

now expresses that the state of $E$ represented by $a$ is valid in the modelled reality during time interval $I$. This denotation is also used to indicate the n+3-tuple $\left(a_{0},\ a_{1},\ a_{2},\ \ldots,\ a_{n},\ s,\ e\right)$. The denotation

\begin{equation}
\label{eq:def-element}
\left(R,\ \left(a,\ \left[s,\ e\right]\right)\right)
\end{equation}

now expresses that the n+3-tuple $\left(a,\ \left[s,\ e\right]\right)$ = $\left(a_{0},\ a_{1},\ a_{2},\ \ldots,\ a_{n},\ s,\ e\right)$ is an element of the extention of $R$.

%\begin{definition}
%Consider a relation $R$ in a temporal database, an entity $A$ given by the attributes $\left(a_1, \ldots, a_n \right)$ and the crisp time interval $I$ given by $\left[a,\ b\right] : a,b \in \T, a \leq b$. The pair $\left(A, I\right)$ expresses that the data regarding the entity A is valid during the crisp time interval I.  Note that the relation $R = \left(A, I \right) x \left(A, I \right)$. The following equation indicates that the pair $\left(A, I\right)$ is in the relation R:
%\begin{equation}
%\label{eq:rel-def}
%\left( R, \left(A , I\right) \right)
%\end{equation}
%\end{definition}

%\begin{example} \textcolor{red}{TODO: INTEGRATE THIS EXAMPLE IN THE `GENERAL EXAMPLE'}
%\label{ex:library-database}
%Consider a library database. Each physical book is stored with an identifier (ID), a title and the start and end loan dates respectively. Table \ref{tbl:library-sample} illustrates the first version of the database, after three insertions. In this example, $A = ($ ID, Title $)$ and $I = ($ Start Loan, End Loan $)$.
%\end{example}

%\vglue13pt
%\begin{table}[htbp]

%In order to simplify the algorithms for the manipulation of data, some auxiliary functions and constants are defined:
In order to simplify the definition of the CRUD functions, some auxiliary functions and constants are first defined.

For the valid-time relation $R$ to be able to represent that the state of entity $E$, represented by $a$, is currently valid, the time domain used by the temporal attributes ($Start$ and $End$) should contain a special constant called \emph{Until Changed} (\emph{UC}).

\begin{definition}
 \label{def:until-changed}
\textbf{Until Changed (\emph{UC}).}\\
Given a time domain $\T$, \emph{Until Changed} is a single value in $\T$ representing an instant infinitely far in the future.
\end{definition}
\vspace{10pt}

This constant is special in that sense that if the value of a tuple of $R$ for attribute $End$ is $UC$, the state, represented by the values of this tuple for the non-temporal attributes, is considered to be valid currently and during an unlimited amount of time in the future.

%To show that the data related to the entity $\left(A, I \right)$ is up-to-date, a special constant called \emph{Until Changed} is defined for the endpoint, $d$, of the time interval $I = \left(c, d\right)$. The value for this constant is considered to be $+\infty$.

\begin{definition}
\label{def:current-tuple}
\textbf{Current Tuple.}\\
The n+1-tuple $a$ is \emph{current} in relation $R$ if and only if:

\begin{equation}
\exists\ c\ \in\ \T|\ \left(R,\ \left(a,\ \left[c,\ UC\right]\right)\right) \nonumber
\end{equation}

\end{definition}

%If the pair $\left(A, I \right)$ with $I = \left(c, UC \right), c \in \T$ is in the relation $R$, then, it is said that the pair $\left(A, I \right)$ is \emph{current} in the relation.

For example, the last row in Table \ref{tbl:library-sample} visualizes the values representing a time interval $\left[4/4/2012 , UC \right]$. The interpretation is that the book with ID = 3 was loaned on 4/4/2012 and it is still loaned. The tuple representing the book with ID = 3 is current in the relation.

The concepts above now allow the introduction of the function \emph{Current}:

\begin{align}
\text{Current} & :\ \Rel \times R_{NT} & \rightarrow & Ex_{R} \nonumber \\
               & :\ \left(R,\ a\right) & \rightarrow &
 \begin{cases}
  \left\{\left(a,\ \left[s,\ e\right]\right)\right\} & \text{if } \exists \left(a,\ \left[s,\ e\right]\right)| \\
  									 & (\left(R,\ \left(a,\ \left[s,\ e\right]\right)\right)\\
  									 & \wedge \left(\left[s,\ e\right]\ =\ \left[c,\ UC\right]\right)), \\
  									 & c \in \T \\
  \emptyset & \text{else}
 \end{cases}
 \nonumber
\end{align}

Here, $Ex_{R}$ denotes the extention of $R$, $\Rel$ denotes the set of all relations in the database schema and $A_{NT}$ denotes the set of all n+1-tuples consisting of the non-temporal values of the tuples of relation $R$.

%\begin{definition}
%\label{def:current-in-relation}
%\textbf{Function `Current'.}\\

%This function returns a crisp time interval $I$ if the entity $A$ is current in the relation $R$ or the empty set, if the entity $A$ is not current in the relation.

%Consider an object $A$ and a relation $R$ in a temporal database. The function Current$\left(R, A \right)$ obtains the tuple $\left(A, I \right)$ that is current in the relation $R$, that is it, $I = \left[a, +\infty \right]$.

%\begin{align}
%\label{eq:current-in-relation}
%\mbox{Current} \left(R, A \right) &=& \\ 
%\begin{cases}
%\nonumber
%I & \mbox{ if } (A,I) \in R \wedge I = \left(c, UC \right), c \in \T\\
%\emptyset & \mbox{ in any other case }
%\end{cases}
%\end{align}
%\end{definition}

For example, the expression Current($R_{ex}$, $\left\{3,`Dracula'\right\}$), where $R_{ex}$ denotes the example relation visualized in table \ref{tbl:library-sample}, returns the tuple $\left\{3, `Dracula', 4/4/2012 , UC\right\}$. On the other hand, Current($R_{ex}$, $\left\{4,`Frankenstein'\right\}$) returns the empty set.

%The following two Allen's relations (see Figure \ref{fig:allen-rel}) are defined~\cite{Nagypal2003} to implement the auxiliary functions.

Given two time intervals $I$, represented by $\left[s_{I},\ e_{I}\right]$, and $J$, represented by $\left[s_{J},\ e_{J}\right]$, the following two functions \emph{During} and \emph{Overlaps} are functions used to compare the time intervals $I$ and $J$, based on sequential comparisons of the values representing their starting and ending instants.

\begin{align}
\text{During} & :\ \Pow(\T)\ \times\ \Pow(\T) & \rightarrow & \bool \nonumber \\
              & :\ \left(\left[s_{I},\ e_{I}\right],\ \left[s_{J},\ e_{J}\right]\right) & \rightarrow & \left(s_{I}\ >\ s_{J}\right)\ \wedge\ \left(e_{I}\ <\ e_{J}\right)
              \nonumber
\end{align}

\begin{align}
\text{Overlaps} & :\ \Pow(\T)\ \times\ \Pow(\T) & \rightarrow & \bool \nonumber \\
							  & :\ \left(\left[s_{I},\ e_{I}\right],\ \left[s_{J},\ e_{J}\right]\right) & \rightarrow & \left(s_{I}\ <\ s_{J}\right)\ \wedge\ \left(e_{I}\ <\ e_{J}\right)\nonumber \\
							  & & & \wedge\ \left(e_{I}\ >\ s_{J}\right)
							  \nonumber
\end{align}

Here, $\Pow(\T)$ is the powerset of $\T$ and $\bool$ is the Boolean domain $\bool\ =\ \left\{T,\ F\right\}$, where $T$ is a value representing the logical constant `True' and $F$ is a value representing the logical constant `False'. Now, the function \emph{CloseR} can be introduced. CloseR is used to `close' the representation of a time interval with respect to the representation of another time interval.

\begin{align}
\text{CloseR} & :\ \Pow(\T)\ \times\ \Pow(\T) & \rightarrow & \Pow(\T) \nonumber \\
							& :\ \left(\left[s_{I},\ e_{I}\right],\ \left[s_{J},\ e_{J}\right]\right) & \rightarrow & 
							\begin{cases}
							\left[s_{I},\ s_{J}\right[ & \text{if } \left(s_{J}\ =\ UC\right)\ \wedge\\
							 & \text{During}(\left[s_{J},\ e_{J}\right], \\
							 & \left[s_{I},\ e_{I}\right]) \\
							\left[s_{I},\ e_{I}\right] & \text{else }
							\end{cases}
							\nonumber
\end{align}

For example, consider two time intervals, represented by $I = \left[4/4/2012, UC \right]$ and $J = \left[24/4/2012, UC \right]$ respectively. Then, CloseR$\left(I, J \right) = \left[4/4/2012, 23/4/2012 \right]$.

@@

The functionality provided by functions Current and CloseR now allows to introduce the necessary CRUD-functions.

%Now it is possible to `close' the current version of an entity by using \eqref{eq:close-a-crisp-interval} and \eqref{eq:current-in-relation}. This functionality is required to append new information about an existing entity in the relation.

%\begin{definition}
%\label{def:close-current-version}
%Consider an entity $A$, the relation $R$ and a time interval $J$. Then, the function close-current$\left(R,\left(A,J\right) \right)$ closes any current version of the object $A$ if it exists and add the new version $\left(A, J \right)$.

%\begin{eqnarray}
%\label{eq:close-current}
%\text{Close-current} \left(R, \left(A, J\right) \right) =\\
%\begin{cases}
%\nonumber
%R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseR } \left(I, J\right) \right)\right \rbrace \cup \left \lbrace\left(A, J\right) \right \rbrace  \\
%\nonumber
%\mbox{ if } I = \mbox{ Current } \left(R, A \right)\\ %\wedge J > \mbox{ CloseR } \left(I, J \right)   \\
%\nonumber R , \text{ in any other case}
%\end{cases}
%\end{eqnarray}
%\end{definition}

%For example, consider the book with ID = 3 in Table \ref{tbl:library-sample}. The current version was loaned on 4/4/2012. The book was returned to the library but, for some reason, the end loan date was not registered. The book is loaned again on 24/4/2012. Then, the Close-current function allows to close the current version of the book and insert a new version. First, the function CloseR is applied with $I = \left(4/4/2012, UC \right)$ and $J = \left(24/4/2012, UC \right)$. Hence, the value for the time interval $I$ is $ \left(4/4/2012, 23/4/2012 \right)$. Then, the modifications on the value for the time interval $I$ are stored. Finally, a new row with the current values of the book and the time interval $J$ is stored.
%The result of this operation is illustrated on Table \ref{tbl:library-sample-2}.
